Performance of prediction models for opportunities for improvment in trauma care across patient cohorts

### 2023-12-17 manuscript.Rmd
Edit and comment on manuscript

output.format <- knitr::opts_knit$get('rmarkdown.pandoc.to')

@@

To enhance the precision of the selection process for mortality and morbidity conferences, there have been efforts to introduce trauma mortality prediction models. Nevertheless, the performance of these models has been unsatisfactory, being poor predictors for OFI [@Ghorbani2018; @Heim_2016]. This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care. Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement [@Attergrim2023].

@@

To assess the difference in performance of prediction models for opportunities for improvement in trauma care across different clinical trauma cohorts.

@@

All variables from the trauma registry are considered potential predictors in our analysis. The models have been developed and trained on the data set previous to this study [@Attergrim2023]. Eight models are included, all being developed using the Tidymodels framework [@tidymodels]. These include logistic regression (LR), random forest (RF), decision tree (DT), support vector machine with a radial basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor (k-NN).

@@

```{r tab1, echo=FALSE, fig.cap="Selected sample characteristics", fig.align='center' }
if (output.format == "latex") {
  tab1 <- t1kable(create.tableone(dataset), "latex", longtable = T, booktabs = T) %>%
    kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "",threeparttable =  TRUE, escape = FALSE)
} else if (output.format == "docx") {
    tab1 <- t1flex(create.tableone(dataset)) %>%
      add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.") %>%
      autofit()
}

@@

Karolinska trauma registry contains 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2). Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.

@@

The biggest cohort in the study was "other cohort" with 6707 (81%) of patients, followed by "blunt multisystem without TBI" at 9% (n=708). They also are the most common cohorts with OFI, making up 64% (n=329) and 23% (n=116) respectively. See table 1 for patient characteristics.

@@

We compared the different cohorts with AUC as our performance metric. (See table 2) Isolated severe TBI had the highest AUC at 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762). The model that performed the lowest in the cohort "Blunt multisystem with TBI" (AUC 0.696, 0.692-0.701) Figure 2 shows details of the receiver operating curve. The P-value between the different cohorts was 1 x 10 \^-6.

@@

We set out to assess machine learning models viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts. We found that there is some variance in the models performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study. The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.

A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by our models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study. Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789). A 2020 study found similarly good performance of machine learning models in TBI patients looking to predict poor outcomes. [@Matsuo2020] This may suggest that some models inclination to perform well in prediction of traumatic brain injury, whereas outcomes linked to other trauma cohorts may be harder to predict.

@@

Secondly, our learners are dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset. As a consequence, the models are prone to some of the same limitations as the current system of audit filters. The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting OFI. This introduces an inherent bias. The audit filter system favour some types of error, while possibly missing others. At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI. Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [@Albaaj2023]. This in turn introduces possible selection bias, giving a skewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.

@@

In this study, we set out to investigate the viability of such models across cohorts. The findings suggest that not all patient types of patient cases perform as well in this context, and could need further improvement before implementation in selection of cases to be discussed at a conference. This is important from a fairness perspective, as the type of injury should not dictate the opportunity to find and correct errors in clinical practise.

>>>>

output.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

@@

To enhance the precision of the selection process for mortality and morbidity conferences, there have been efforts to introduce trauma mortality prediction models. Nevertheless, the performance of these models has been unsatisfactory, being poor predictors for OFI [@Ghorbani2018; @Heim_2016]. This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care. Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement, and the learner XGBoost showed particular promise [@Attergrim2023].

@@

To assess the difference in performance of a machine learning based prediction model for opportunities for improvement in trauma care across different clinical trauma cohorts.

@@

All variables from the trauma registry are considered potential predictors in our analysis. The models have been developed and trained on the data set previous to this study [@Attergrim2023]. Eight models were included in the original study, all being developed using the Tidymodels framework [@tidymodels]. These include logistic regression (LR), random forest (RF), decision tree (DT), support vector machine with a radial basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor (k-NN). In this study, we focused on the model developed using XGBoost, as this model performed the best in the original paper.

@@

```{r tab1, echo=FALSE, fig.cap="Selected sample characteristics", fig.align='center' }
if (output.format == "latex") {
  tab1 <- t1kable(create.tableone(dataset), "latex", longtable = T, booktabs = T) %>%
    kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "", threeparttable = TRUE, escape = FALSE)
} else if (output.format == "docx") {
  tab1 <- t1flex(create.tableone(dataset)) %>%
    add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.") %>%
    autofit()
}

@@

The Karolinska trauma registry includes 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2). Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.

@@

The biggest cohort in the study was "other cohort" with 6707 (81%) of patients, followed by "blunt multisystem without TBI" at 9% (n=708). These were also the cohorts with the most OFI, making up 64% (n=329) and 23% (n=116) respectively. See table 1 for patient characteristics.

@@

We compared the performance of the ML model in the different cohorts with AUC as our performance metric (See table 2). The ML model performed best in the cohort of patients with isolated severe TBI, in which it had an AUC of 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762). The ML model performed worst in the cohort of patients with blunt multisystem trauma with TBI (AUC 0.696, 0.692-0.701). Figure 2 shows details of the receiver operating curves. The P-value between the different cohorts was 1 x 10 \^-6, meaning that there were significant differences in AUC between cohorts.

@@

We set out to assess the performance of a model based on the machine learner XGBoost's viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts. We found that there is some variation in the model's performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study. The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.

A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by ML models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study. Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789). A 2020 study found similarly good performance of machine learning models in TBI patients looking to predict poor outcomes. [@Matsuo2020] This may suggest that some models inclination to perform well in prediction of traumatic brain injury, whereas outcomes linked to other trauma cohorts may be harder to predict.

@@

Secondly, our learner is dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset. As a consequence, the models are prone to some of the same limitations as the current system of audit filters. The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting<!-- Do you mean determining?--> OFI. This introduces an inherent bias. The audit filter system favour some types of error, while possibly missing others. At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI. Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [@Albaaj2023]. This in turn introduces possible selection bias, giving a skewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.

@@

In this study, we set out to investigate the viability of a model based on the machine learner XGBoost across cohorts. The findings suggest that the model does not perform as well across all patient types, and could need further improvement before implementation in selection of cases to be discussed at a conference. This is important from a fairness perspective, as the type of injury should not dictate the opportunity to find and correct errors in clinical practise.
